The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2016-03-19 script/en/chapters/primes.tex
CT Book ed 12: Added 3 graphics in primes.tex -- at the beginning and into section sieve of Eratosthenes. Started to search for a old /index problem [so later when solved the error detecting patterns ("\index{ZZZ") have to be taken off again.]

\begin{theorem}\label{thm-pz-sqr}
Each whole number $m$ greater than $1$ possesses a lowest factor greater than
$1$. This is a prime number $p$. Unless $m$ is a prime number itself, then: $p$
is less than or equal to the square root of $m$.
\end{theorem}

@@

An easy way of calculating all $PI(x)$ primes less than or equal to $x$ is to
use the sieve of Eratosthenes. In the 3rd century B.C., he found an
extremely easy, automatic way of finding this out. To begin with, you write
down all numbers from 2 to $x$, circle 2, then cross out all multiples of 2.
Next, you circle the lowest number that hasn't been circled or crossed out
(3) and again cross out all multiples of this number, etc. You only need to
continue until you reach the largest number whose square is less than or
equal to $x$.%
\footnote{%
    \index{NT, Learning Tool for Number Theory}%
    \index{educational tool NT}%
    With the educational tool for number theory {\bf NT} you can apply the
    sieve of Eratosthenes in a computer-aided and guided way: Enter you own
    number and do the sieving step by step:
    See the NT learning unit 1.2, pages 6/21 and 7/21.\\
    NT can be called in CT1\index{CrypTool 1} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix \ref{s:appendix-Learn-NT}.\\
    CT2\index{CrypTool 2} contains a visualisation of this method within
    the tutorial \textbf{``World of Primes''}.
}

>>>>

\newpage
The following table gives a first impression how primes are distributed between natural numbers. There are many graphical forms of represenation
%display formats
(the most well-known is the Ulam spiral). However till now, these graphical forms gained no new insights, but for some people they created the impression that
%are suggestive of
there are at least local patterns within the random distribution.

\begin{figure}[!hb]
\begin{center}
\includegraphics[scale=0.3]{figures/Numberrectangle-with-colored-primes_Irisprime.jpg}
\caption[Primes between the first 390 integers -- marked with color]
        {Primes between the first 390 integers -- marked with color\protect\footnotemark}
\label{Primes-in-a-390-integer-rechtangle-figure}
\end{center}
\end{figure}
\footnotetext{%
   Graphics from 
   \url{http://mathforum.org/mathimages/index.php/Image:Irisprime.jpg}, 30*13 rectangle
   }

\begin{figure}[!hb]
\begin{center}
\includegraphics[scale=0.41]{figures/Ulam_spiral_via-CT2-within-1-999.png}
\caption[Primes between the first 999 integers -- as Ulam spiral]
        {Primes between the first 999 integers -- as Ulam spiral\protect\footnotemark}
\label{Primes-in-a-999-integer-ulam-spiral-figure}
\end{center}
\end{figure}
\footnotetext{%
   Graphics from CT2, menu Crypto Tutorials, Distribution of primes, Ulam's spiral; 32*32 points
   }
%\newpage
\begin{figure}[!hb]
\begin{center}
\includegraphics{figures/Ulam_spiral_Wikipedia.png}
\caption[Primes between the first 4000 integers -- as Ulam spiral]
        {Primes between the first 4000 integers -- as Ulam spiral\protect\footnotemark}
\label{Primes-in-a-4000-integer-ulam-spiral-figure}
\end{center}
\end{figure}
\footnotetext{%
   Graphics from 
   \url{http://mathforum.org/mathimages/index.php/Image:Ulam_spiral.png}, 200*200 Ulam spiral
   }

\newpage
\begin{theorem}\label{thm-pz-sqr}
Each whole number $m$ greater than $1$ possesses a lowest factor greater than
$1$. This is a prime number $p$. Unless $m$ is a prime number itself, then: $p$
is less than or equal to the square root of $m$.
\end{theorem}

@@

An easy way of calculating all $PI(x)$ primes less than or equal to $x$ is to
use the sieve of Eratosthenes. In the 3rd century B.C., he found an
extremely easy, automatic way of finding this out. To begin with, you write
down all numbers from 2 to $x$, circle 2, then cross out all multiples of 2.
Next, you circle the lowest number that hasn't been circled or crossed out
(now 3) and again cross out all multiples of this number, etc.

\begin{figure}[!hb]
\begin{center}
\fboxrule0.5pt \fbox{
 \includegraphics{figures/Primes-via-Eratosthenes-from-Wikipedia_en.png}
 }
 \caption[The sieve of Eratosthenes, applied to the first 120 numbers]
         {The sieve of Eratosthenes, applied to the first 120 numbers\protect\footnotemark}
\label{SieveEratosthenes01-figure}
\end{center}
\end{figure}
\footnotetext{%
   Graphics from 
   \url{https://upload.wikimedia.org/wikipedia/commons/0/0b/Sieve_of_Eratosthenes_animation.svg}
   }

You need to continue only until you reach the largest number whose
square is less than or equal to $x$ (here up to 10, as $11^2$ is already $> 120$).%
\footnote{%
    \index{NT, Learning Tool for Number Theory}%
    \index{educational tool NT}%
    With the educational tool for number theory {\bf NT} you can apply the
    sieve of Eratosthenes in a computer-aided and guided way: Enter you own
    number and do the sieving step by step:
    See the NT learning unit 1.2, pages 6/21 and 7/21.\\
    NT can be called in CT1\index{CrypTool 1} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix \ref{s:appendix-Learn-NT}.\\
    CT2\index{CrypTool 2} contains a visualisation of this method within
    the tutorial \textbf{``World of Primes''}.
}
